The Toki Pona Language: an overview and some hacks

### 2017-11-11 article/article.tex
first draft almost done

Other particles are: a or kin, o, taso, anu, en, 
nanpa, seme. mu?,

Vocatives should have an o after the noun phase.

@@

toki e ni, wile e ni, etc.
'e ni' can be omitted and : used by itself.

Names are by default transliterated,

@@

\section{Analysis and hacks for the language}\label{hacks}
% number of possible syllables
% number of possible words of size X (1, 2, 3, 4....).
% syntax highlighting, change colors on the fly
% analysis of the language
% synthesis of sentences and texts
% hacks from other ppl
\subsection{Hacks from other people}
The \url{tokipona.net} has a number of tools,
just as to transliterate names into Toki Pona phonemes,
and search in corpus.

\section{Conclusions and further work}\label{conc}
\begin{itemize}
  \item Relate Toki Pona to Wordnet: should one Toki Pona word
    be related to more then one synset of the English language?
  \item Understand how the corpus is gathered in \url{tokipona.net}.
  \item Know about previously existing words that were used for Toki Pona
    (e.g. suno and suwi might come from sun and sweet),
    and about the reasons that lead Sonja (and maybe other people)
    to choose the 14 letters and the syllable structure.
    This might require a dedicated communication with the
    speaker community and the documentation authors.
  \item Corpus-based analysis.
  \item Publication of original texts and translations.
\end{itemize}
\subsection*{Acknowledgments}
FAPESP (project 2017/05838-3); Vim developers and documentation maintainers;
Vim user community.

@@

>>>>

Other particles are:
\begin{itemize}
  \item a or kin, emphasis.
  \item o, vocative or imperative ('jan lukin sitelen o, li wawa')
  \item taso, means however or 'only' if adjective.
  \item anu, en: 'or' and 'and'. Used for nouns in nous phrases.
    For repetition of verbs, repeat li.
    For object nouns, repeat e.
    If the noun is complementing a preposition (tawa, lon),
    one might repeat the preposition.
    As Toki Pona is a recent language, and is able to cope with
    variation due to its simplicity, I would advocate for
    using en and anu wherever there is no ambiguity.
  \item nanpa, denotes numbering.
  \item seme, for questions, used next to the thing being asked for.
    'Why?' might be expressed as
    'seme la sina pana e moku lon sewi'.
  \item mu, for animal noises. For me it is not a particle, as in the
    official dictionary, but a noun.
    I also like to use it as a verb:
    'mi pakala e luka. mu mute.
\end{itemize}

The vocabulary specifies morphosyntactic classes:
nouns, adjectives, verbs, pre-verbs, adverbs, particles, prepositions, and numbers.
I find that they might help the user and newcomer, but
it might also suggest a deviation from what I understand and read:
the words might be used indistinctly to be the nouns
(subjects, the predicate when there is no object, objects,
and preposition complements),
the adjectives (anything that does not start the noun phrase or follows a pi),
verbs
(follows mi or sina or li or a preposition),
adverbs (follows the verb).
The pre-verbs (wile, ken, awen, kama, lukin, sona),
might follow a verb, but might also be understood
as the verb qualified with the next word,
which carries a very similar if not identical meaning.
The pre-verbs are all also defined as other morphosyntactic classes,
such as adjective, noun, verb.
The only exception is wile, which is only a pre-verb.

Thus, the classes given in the dictionary dictate little
in practice:
jan kala li lape lon ni.
Where kala, lape and ni are in this phrase
as adjective, verb and noun,
and are in the dictionary as noun,
adjective and adjective.

As far as I can see, one should regard
the particles li, e, pi, and la
and punctuation.
The other tokens of the vocabulary
might be used in any of the remaining positions.
Detection:
\begin{itemize}
  \item Noun: the first word in a noun phrase.
    After an 'e' and after a pi,
    The first word in the sentence if
    sentence does not start by the verb.
    Might be in the position of the verb
    if the sentence has no object.
  \item Adjective: second word on after an e and after
    Second word on in the subject phrase if present.
  \item Verb: after a li, mi or sina.
    If there is no object, the verb position
    is often a noun.
  \item After a preposition,
    there can be a noun phrase, a verb phrase
    or nothing.
  \item Notice that there is ambiguity in the structure
    introduced by the omission of li after mi and sina.
    Also, when there is no object, a noun or a verb
    or an adjective might be in the verb position
    if there is no object.
    The prepositional complement is also not defined.
    (mi moku tawa pali, tawa tomo).
    So, these are sources of syntactic ambiguities
    in Toki Pona.
    They might be solved or minimized by using the semantics
    of the words.
    One preliminary effort in this direction might be
    using the classes in the official dictionary to resolve
    ambiguities whenever possible.
    This solution is not optimal in correct POS tagging,
    and does not solve all possible ambiguities
    (there are words classified as nouns and adjectives,
    adjectives and verbs, nouns and verbs, particle and verb).

    Another source of ambiguity is the pre-verbs as described
    in the literature~\cite{tpLang,pije}.
    But I find it reasonable to understand them as verbs.

\end{itemize}

Also, the prepositional complement might
be a noun or a verb.
I could not come up with a sentence
where it would be understood as an adjective.

@@

In formations such as
toki e ni:, wile e ni:, tan ni: etc.
'(e) ni' can be omitted and : used alone.

Names are by default transliterated,
but might not be, as described in Section~\ref{mytoki}.

@@

\section{Analysis and hacks for the language}\label{hacks}

\section{Statistics of the vocabulary}
In~\cite{pije} there is statistics about Toki Pona corpus.
This section focuses on the statistics of the vocabulary
and syntactic rules:
the letters, phonemes, word sizes,
possible combinations for words and sentences.
The statistics are in Appendix~\cite{listings},
and the next paragraph is an overview.
Python scripts were used to obtain the measurements
and are available at~\cite{tokipona}.

As described in Section~\cite{basics},
there is only 14 letters,
and phonemes respect a few rules.
There are 120 different words in the official vocabulary,
4 of them having synonyms.
A total of 124 tokens.
Not counting proper nouns (names)
and punctuation.
They include X nouns, Y verbs, Z adjectives,
W prepositions, Y particles, K pre-verbs.
Of which one might distinguish only between
the particles and the other J words that
might be used anywhere a particle will not.

Most often vowels,
consonants, consonant vs vowel.
Most often phonemes in general an specific positions..
Possible phonemes given the rules.
Comparison of the occurring against the possible.

Syntactic structures:
Possibilities on noun, verb and prepositional phrases
and in sentences.
Counting along the number of words.

\section{Synthesis of text}
The same package~\cite{tokipona}
has capabilities for synthesizing text.
Noun, verb and prepositional phrases,
sentences.
It also aims at making larger scale texts
by keeping a record of the used words and structures (context)
and using stylistic outlines for poems and short narratives.

% number of possible syllables
% number of possible words of size X (1, 2, 3, 4....).

\section{Syntax highlighting}
The same package~\cite{tokipona}
has a Vim syntax highlighting plugin
for Toki Pona.
Instructions for installing and using
the syntax highlighting is at~\cite{tokipona}.

Basically, it distinguishes the words among the morphosyntactic
classes according to the official dictionary.
As a word often belongs to more than one class,
The precedence of them might be set by the user.
Also, some classes might be further refined or joined,
such as by distinguishing only particles and the rest,
or maybe particles and prepositions and the rest.
The colors are also promptly changed according to~\cite{vimArt}
and exemplified in the package documentation.

Currently, the Python package synthesizes the
syntax file.
The user has control of class precedence and
merging.
The choice of precise coloring schemes
might involve hacking the colorscheme being
used in Vim (such as 'blue', 'elflord' and 'gruvbox'),
and Vim's highlighting schemes as described in~\cite{vimArt}.
In summary,
the usage of the package and plugin might be performed
through the following actions:
\begin{itemize}
  \item Installation of the plugin.
  \item Tweak of the syntax file by hand.
  \item Running the Python script to generate a new syntax file
    according to other settings.
  \item Write a file inside Vim using Toki Pona and save
    the file with the .tokipona extension.
    Reload the highlighting scheme whenever you
    change the syntax file by hand or through the Python script.
  \item Access the used highlighted groups with :syntax,
    Access all the highlighting groups with :so \$VIMRUNTIME/syntax/hitest.vim.
    Change the coloring of a set of terms by associating
    a used group (e.g. tokiponaADJECTIVE) to an existing group (e.g. Visual):
    :highlight link tokiponaADJECTIVE Visual.
    The plugin comes with the :TokiStation command, which opens a window with
    the files: ijositelen.tokipona, makevimSyntax.py, Highlight Test (created by hitest.vim above), syntax/tokipona.vim.
    Another tab with shells: python to run over the makeVimSyntax.py and make new syntax/tokipona.vim files. Another with Readme, and PDF documentation. Another with an IPython.
    Imported tokipona. It resets completely upon command :TokiClose, closing all created windows.
\end{itemize}

\subsection{Hacks from other people}
The \url{tokipona.net} has a number of tools,
just as to transliterate names into Toki Pona phonemes,
and search in corpus.

\section{Conclusions and further work}\label{conc}
\begin{itemize}
  \item Relate Toki Pona to Wordnet: should one Toki Pona word
    be related to more then one synset of the English language?
  \item Understand how the corpus is gathered in \url{tokipona.net}.
  \item Know about previously existing words that were used for Toki Pona
    (e.g. suno and suwi might come from sun and sweet),
    and about the reasons that lead Sonja (and maybe other people)
    to choose the 14 letters and the syllable structure.
    This might require a dedicated communication with the
    speaker community and the documentation authors.
  \item Corpus-based analysis.
  \item Publication of original texts and translations.
  \item Make an article written in Toki Pona.
    I wrote Section~\ref{finalToki}, and believe that
    a summary in both English and Toki Pona for
    facilitating the acquisition of context,
    a reasonable article on some scientific topics is
    possible.
    I first conceived something around complexity, statistics, physics, or computer science.
    But one possibility is to write about linguistics, philosophy, literature
    or psychology with the partners I write in English.
    I can start a draft,
    they might learn the language in a few hours (with or without my help),
    to contribute, and we can write a short paper.
  \item Enhance the synthesis of text to yield better contextualized text
    and stylistic traces, such as for poems and short stories.
    Give the user the ability to choose the sentences (generate randomly according to
    previously written or given text,
    some rules input by the user, the package and the language guidelines,
    outputs to the screen and asks to keep and discard).
  \item Enhance the syntax highlighting implementation described in Section~\ref{high}.
    It uses only syntactic cues to choose POS tags,
    which might be overcome by using n-grams and
    further techniques from Natural Language Processing~\cite{POS}.
\end{itemize}

\subsection*{Acknowledgments}
FAPESP (project 2017/05838-3); Vim developers and documentation maintainers;
Vim user community.

@@

Proposed notations for numbers seem numerous.
I tend to think that one might indicate is two numbers
are multiplying (pi) or are in different scales
(such as in decimal or binary notations).
For example, I take luka two to mean 52.
Or it might not be taken to such level of strictidness,
for 52 is a reasonable notation for a simple language.
mi jo e jan sama nanpa 12.
Or even:
ona li lon e soweli 27.

Remove the 'li'.
I've been avoiding e ni: and using only :.
've been omitting the subject if it is the
same as in the last sentence.
I've been avoiding also the li sometimes,
and starting only a predicate + object phrase,
or a whole phrase altogether.

\input{pos}

\section{Final words in Toki Pona}
toki li nasin e lawa.
li nasin tawa (pi) toki insa.

% jan li ken toki li ken pali pona e toki pona.
toki pona li pona e nasin tan ni:
ona li jo e sitelen mute lili.
ona li pona.
pona kepeken weka 'p' li ona, a.

o taso la toki pona li kalama li lukin pona.
sitelen en nasin li open e sitelen suli\cite{tpLang,pije,fb1,fb2,tokisona,Wikipesija}.
li sona. li nasin e toki e sona e lawa e lon.

toki ni li wawa tawa jan mute nasin en toki.
wawa tawa toki pi jan sona.
pi ilo nanpa en nanpa nasin.
taso tawa toki sona a.
sitelen sona, sitelen musi.

ilo lon linha~\ref{hacks} li pana e sitelen
e sona tan sitelen,
en nasin.
linha~\ref{basics} li pana e nasin pi toki pona.
e sitelen tawa kama sona.

mi wile pali e sitelen lon toki pona.
sitelen lon nasin, sitelen sona,
sitelen musi.
ante e toki pona la
la ona li nasa.
taso nasa li pona mute.
li pona mute tawa lawa,
tawa kama sona, tawa sitelen e toki.
ante la toki pona o.
toki e toki pona tawa sina.
kama sona e toki pona sina.

o pona tawa jan pi toki pona.
jan Sonja, Birns-Sprage, Kipo, Pije,
Siwejo, Malija, jan kulupu mute.
